---
output:
  word_document: default
  pdf_document: default
  html_document: default
---
# Demo

This is a step by step demo analyzing FeNO data using our U-HB method and other unified or two-stage methods.

<br>

### 1. Generate dateset for simulation


First, we need to build a dataset which mimics the real FeNO data. Here we use Scenario 1 for example (all NO parameters are affected by environment factor X) There are several key parameters here: number of subjects in the study, number of maneuvers, maneuvers’ flow rates, etc. We set them the same as in the paper but shrink the number of subjects to 20 to obtain a fast example. This number may be expanded to any sample size to see if it is large enough to achieve converged results. We will build a balanced dataset (subjects have same number of maneuvers of each tested exhale flow rate) first and there is also an unbalanced dataset example in section #.


```{r,message=FALSE,eval=FALSE}
rm(list=ls())
demoseed <- 2020
Ndat <- 20
source("/Users/wengjingying/Downloads/FeNO/eno/Demofunctions.R")

# Total number of subjects in simulated datasets


enodata<-Obs_FB(alpha=alpha,beta=beta,X=X,Flow=flow,SD=sderror,NOcov=CovMat_chs_Caw)
save(enodata, file="data_S1.Rdata")
```

<br>

### 2. Model fitting

### 2.1 U-HB

We start fitting U-HB model by using short iterations for a pilot run to check for any existing errors. The initial burning and MCMC iterations (N.iterT/2) were set to be 120 and 120. We let the model go through a 120 iteration update (addon.iter=120, Max_update=1) and process to the 600 iteration posterior sampling stage (n.final=600). Those interations are so short that we expect large Rhats value, which indicate failure of converge. So we also set the Rhat criteria (rhat) as large as 5 for the criteria test. Those iteration numbers should be at least ten times as what we set here to achieve convergence (Rhat<1.1, N.iterT=12000,addon.iter=6000,Max_update=40,n.final=6000).

After setting the parameters, we record the starting time, run a burning process and an initial update, then we check Rhats first, if are less than the  criteria we set, we go to theposterior sampling stage, if not, we update the MCMC up to a maximum times we set to fulfill the Rhat criteria and go to posteria sampling stage.


```{r,eval=FALSE}

# set Gelman Rubin criteria
rhat=5
# set how many iterations needed for update MCMC
addon.iter=120
# maximum updates
Max_update=1
# iterarions needed to generate posterior distribution
n.final=600
# starting iteration numbers (1/3 will be burning)
N.iterT=240
# measurement error
sigma2=1e-3

# generate random initial values
set.seed(demoseed)

inits_FcovC <- 
    list(
      list(
        tau_c=runif(1,0.05,400), #could be unif too
        alpha_Ca=runif(1,0,5),
        alpha_logCaw=runif(1,0,5),
        alpha_logDaw=runif(1,0,5),
        beta_Ca=runif(1,-1,1),
        beta_logCaw=runif(1,-1,1),
        beta_logDaw=runif(1,-1,1)
      ),
      list(
        tau_c=runif(1,0.05,400),
        alpha_Ca=runif(1,0,5),
        alpha_logCaw=runif(1,0,5),
        alpha_logDaw=runif(1,0,5),
        beta_Ca=runif(1,-1,1),
        beta_logCaw=runif(1,-1,1),
        beta_logDaw=runif(1,-1,1)
      ),
      list(
        tau_c=runif(1,0.05,400),
        alpha_Ca=runif(1,0,5),
        alpha_logCaw=runif(1,0,5),
        alpha_logDaw=runif(1,0,5),
        beta_Ca=runif(1,-1,1),
        beta_logCaw=runif(1,-1,1),
        beta_logDaw=runif(1,-1,1)
      )
    )

# set monitors for parameters we want to track
param.R2jags_NOTX <-c("alpha_Ca","alpha_logCaw","alpha_logDaw",
                        "beta_Ca","beta_logCaw","beta_logDaw",
                        "Ca","logCaw","logDaw",
                        "sdCa","sdlogCaw","sdlogDaw","corlogCawCa","corlogCawlogDaw","corlogDawCa",
                        "sigma_c")



# set priors
alphaC_prior=c(2,log(68),log(12))
betaC_prior =c(0,0,0)

flow <- c(rep(30,2),rep(50,2),rep(100,2),rep(300,2)) #specify flow rates

```


```{r,eval=FALSE}
# set data set 
data_cov <- list(Ndat=length(enodata$X),Nflows=length(flow),flow=flow,logeno=enodata$logeno,X=enodata$X,
                 betaC_prior=betaC_prior,alphaC_prior=alphaC_prior)


set.seed(demoseed)

time<-proc.time()[3]
#### truncated model

# Run jags
fit.R2jags_NOXT<- jags(data=data_cov, parameters.to.save=param.R2jags_NOTX, inits=inits_FcovC,
                         model.file=truncationNOX, n.chains=3, n.iter=N.iterT, n.burnin=N.iterT/2) 
N_updateT=0 #count how many updates needed
N_final=0 # count how many times the first check was met
while(N_updateT<Max_update){
  # check weather the Rhats of intereset parameters was less then pre-set criteria
    if(max(fit.R2jags_NOXT$BUGSoutput$summary[rownames(fit.R2jags_NOXT$BUGSoutput$summary) %in% param.R2jags_NOTX,"Rhat" ])<rhat){
      # if yes
      fit.R2jags_NOXT<-update(fit.R2jags_NOXT,n.iter=n.final)
      N_final=N_final+1
      # print the Rhat after the final runs to see if it is stable
      print(fit.R2jags_NOXT$BUGSoutput$summary[rownames(fit.R2jags_NOXT$BUGSoutput$summary) %in% param.CHS_Xs,"Rhat" ])
      # check Rhat again
      if(max(fit.R2jags_NOXT$BUGSoutput$summary[rownames(fit.R2jags_NOXT$BUGSoutput$summary) %in% param.R2jags_NOTX,"Rhat" ])<rhat){
        # if yes we are done, if not, we back to update step again
        break
      }
    }
    fit.R2jags_NOXT<-update(fit.R2jags_NOXT,n.iter=addon.iter)
    N_updateT<-N_updateT+1
    print("update")
    print(N_updateT)
  }


# Thats how we know how long we needed for U-HB
timespend<-proc.time()[3]-time
print(timespend)
# we also save summary informations for trimed mean value and quantiles other than defauld Jags summary (mean, Median, SD, etc)
trimmean<-apply(fit.R2jags_NOXT$BUGSoutput$sims.matrix,2,function(x) mean(x, 0.2))
quantile1_99<-t(apply(fit.R2jags_NOXT$BUGSoutput$sims.matrix,2,function(x) quantile(x, c(0.005,0.995))))
result<-list(
  DIC=fit.R2jags_NOXT$BUGSoutput$DIC,
  pD=fit.R2jags_NOXT$BUGSoutput$pD,
  timespend=timespend,
  summary=cbind(fit.R2jags_NOXT$BUGSoutput$summary,"trimmean"=trimmean,quantile1_99),
  Iter_T=N.iterT+N_updateT*addon.iter+n.final*N_final,#total not summary iter
  Iter_track=c(N.iterT=N.iterT,N_updateT=N_updateT,addon.iter=addon.iter,n.final=n.final,N_final=N_final)
)

fit.R2jags_NOXT$BUGSoutput$summary[rownames(fit.R2jags_NOXT$BUGSoutput$summary) %in% param.R2jags_NOTX,"Rhat" ]

save(result, file="UHB_S1.Rdata")

```

<br>


### 2.2 TS-NLS

NLS uses a different dataset structure from the one used for U-HB. It fits parameters for each subjects independently. 
```{r,eval=FALSE}
logenodata<-melt(data.frame(id=c(1:Ndat),enodata$logeno),id="id")
logenodata<-logenodata[order(logenodata$id,decreasing = FALSE),]
datOut <- data.frame(id=rep(c(1:Ndat),each=length(flow)),eno=NA,logeno=NA,flow=rep(flow,Ndat),X=rep(enodata$X,each=length(flow)))
datOut$logeno <- logenodata$value
datOut$eno    <-exp(datOut$logeno)
datOut$invFlow<-1/datOut$flow
datOut$enoflow<-datOut$flow*datOut$eno

dat <- datOut
X <- enodata$X

starttime=proc.time()[3]
# prelminary function for NLS 2-stage and for NLME functions

# fit for TS-NLS
fitnls <- nlsListStart(dat)
# summary how many subjects' data failed to fit NLS
nls_NA<-sum(is.na(coef(fitnls)[,1]))


# generate NLS result
NLSout <- tryCatch(nls_FINAL(fitnls,X), error = function(x) {
    print("Error with nls");
    warnings();
    data.frame(    est=rep(NA,6),
                   lb=rep(NA,6),
                   ub=rep(NA,6),
                   se=rep(NA,6),
                   p=rep(NA,6)
    )
  }
  )

elapsed=proc.time()[3]-starttime
colnames(NLSout)<-paste("NLS",c("est","lb","ub","se","p"),sep="_")
save(elapsed,NLSout,nls_NA, file="NLS_S1.Rdata")
```


### 2.3 TS-HMA

TS-HMA used the same dataset structure as TS-NLS.
```{r,echo=FALSE}
Hogman3rdOrderAlg <- function(fL,fM,fH,eL,eM,eH){  
  # linT approximation on high flows to get S=CaNO and I
  CaNO_est <- S <- (fH*eH - fM*eM)/(fH-fM)
  # solve for I using linT model at medium flow rate (flow=100)
  I <- (fM*eM - S*fM)
  JawNO_est  <- I
  
  # first order approximation starting value
  Dw0_fo <- I/(eL - S)
  
  # iterative algorithm function, 1st order
  italg <- function(SV,nIterAlg=10){
    # object to store Dw values for each iteration
    outIterAlg <- numeric(nIterAlg)
    # set starting value (SV)
    Dw_n <- SV
    for(i in 1:nIterAlg){
      Dw_nplus1 <- outIterAlg[i] <-  -I*(exp(-Dw_n/fL) - exp(-Dw_n/fM))/(eL-eM) # Inf-NAN-error
      diffDw <- abs(Dw_n - Dw_nplus1)
      #print(i);print(Dw_nplus1)
      # stop if values of Dw_n and Dw_nplus1 are close enough
      if(diffDw <0.001){break}
      Dw_n <- Dw_nplus1  
    }
    out <- list(DawNO=Dw_nplus1,nIter=i,diffDaw=diffDw)
    out
  }
  DawNO_result_fo <- italg(Dw0_fo)
  DawNO_est_fo <- DawNO_result_fo$DawNO
  
  CawNO_est_fo <- (I + S*DawNO_est_fo)/DawNO_est_fo
  
  Ic <- I/(1-(fM+fH)*DawNO_est_fo/(2*(fM*fH)+(fH^3-fM^3)/(6*fM^2*fH^2*(fH-fM)/DawNO_est_fo^2)))
  Sc <- S-DawNO_est_fo*Ic/2/fM/fH+(fM+fH)*DawNO_est_fo^2*Ic/6/fM^2/fH^2
  
  # iterative algorithm function, 3rd order
  italgto <- function(SV,nIterAlg=10){
    # object to store Dw values for each iteration
    outIterAlg <- numeric(nIterAlg)
    # set starting value (SV)
    Dw_n <- SV
    for(i in 1:nIterAlg){
      Dw_nplus1 <- outIterAlg[i] <-  -Ic*(exp(-Dw_n/fL) - exp(-Dw_n/fM))/(eL-eM)
      diffDw <- abs(Dw_n - Dw_nplus1)
      #print(i);print(Dw_nplus1)
      # stop if values of Dw_n and Dw_nplus1 are close enough
      if(diffDw <0.001){break}
      Dw_n <- Dw_nplus1  
    }
    out <- list(DawNO=Dw_nplus1,nIter=i,diffDaw=diffDw)
    out
  }
  DawNO_result_to <- italgto(DawNO_est_fo)
  DawNO_est_to <- DawNO_result_to$DawNO
  
  DawNO_est_final <- DawNO_est_to  
  CawNO_est_final <- Ic/DawNO_est_final + Sc
  CaNO_est_final <- Sc
  JawNO_est_final <- DawNO_est_final*(CawNO_est_final - CaNO_est_final)
  
  outfinal <- c(  CaNO=unname(CaNO_est_final),
                  DawNO=unname(DawNO_est_final),
                  JawNO=unname(JawNO_est_final),
                  CawNO=unname(CawNO_est_final)
  )
  # data checks (both should be true)
  ## test CaNO estimate is positive
  test1 <- CaNO_est_final > 0
  ## test whether measured data is mathematically consistent with model
  LHS <- (eL-eM)/(eM-eH)
  RHS <- fH/fL*((fM-fL)/(fH-fM))
  test2 <- LHS < RHS
  ## To change the function to only give non-missing values for 
  ## valid datasets that produce positive CaNO estimates, remove the 
  ## comments from the following two lines
  #if(!(test2)) outfinal[1:4] <- NA # leave in negative CaNO estimates
  #if(!(test1 & test2)) outfinal[1:4] <- NA
  c(outfinal,valid=unname(test2))
}

HMA_FINAL <- function(dat,X){
  ids <- unique(dat$id)
  # Stage I: HMA fits for each person
  estCoef <- matrix(NA,ncol=5,nrow=length(ids))
  rownames(estCoef) <- ids
  colnames(estCoef) <- c("Ca","Daw","Jaw","Caw","valid")
  for(i in 1:length(ids)){
    dati <- subset(dat, id==ids[i], select=c(id,flow,eno))
    dati <- subset(dati[order(dati$flow),],flow!=50) # drop flow=50 since HMA uses 3 flow rates (L/M/H)
    avgFeNO <- tapply(dati$eno,dati$flow,mean,na.rm=TRUE)
    fL <- 30
    fM <- 100
    fH <- 300
    eL <- avgFeNO["30"]
    eM <- avgFeNO["100"]
    eH <- avgFeNO["300"]
    # only run Hogman Algorithm if all 3 (low=30,medium=100,high=300) flows are available
    if(any(is.na(c(fL,fM,fH,eL,eM,eH)))){
      HogmanResultsi <- c(Ca=NA,Daw=NA,Jaw=NA,Caw=NA,valid=NA)
    }
    else{
      HogmanResultsi <- Hogman3rdOrderAlg(fL,fM,fH,eL,eM,eH)
      # if algorithm produces any NaN/Inf, replace with all missing (happens 1x in CHS)
      if(any(is.na(HogmanResultsi))){HogmanResultsi <- c(CaNO=NA,DawNO=NA,JawNO=NA,CawNO=NA,valid=NA)}
    }
    estCoef[i,] <- HogmanResultsi
  }
  estCoef <- as.data.frame(estCoef)
  estCoef$logCaw <- log(estCoef$Caw)
  estCoef$logDaw <- log(estCoef$Daw)
  # Stage II: simple linear regressions
  fit_Ca     <- lm(estCoef$Ca~X)
  fit_logCaw <- lm(estCoef$logCaw~X)
  fit_logDaw <- lm(estCoef$logDaw~X)
  # output
  out <- matrix(NA,nrow=6,ncol=5)
  rownames(out) <- c("Ca.(Intercept)","Ca.X","logCaw.(Intercept)","logCaw.X","logDaw.(Intercept)","logDaw.X")
  colnames(out) <- c("est","lb","ub","se","p")
  out <- as.data.frame(out)
  
  out[c("Ca.(Intercept)","Ca.X"),c("est","se","p")] <- summary(fit_Ca)$coef[,c("Estimate","Std. Error","Pr(>|t|)")]
  out[c("logCaw.(Intercept)","logCaw.X"),c("est","se","p")] <- summary(fit_logCaw)$coef[,c("Estimate","Std. Error","Pr(>|t|)")]
  out[c("logDaw.(Intercept)","logDaw.X"),c("est","se","p")] <- summary(fit_logDaw)$coef[,c("Estimate","Std. Error","Pr(>|t|)")]
  
  out[c("Ca.(Intercept)","Ca.X"),c("lb","ub")] <- confint(fit_Ca)[,c("2.5 %","97.5 %")]
  out[c("logCaw.(Intercept)","logCaw.X"),c("lb","ub")] <- confint(fit_logCaw)[,c("2.5 %","97.5 %")]
  out[c("logDaw.(Intercept)","logDaw.X"),c("lb","ub")] <- confint(fit_logDaw)[,c("2.5 %","97.5 %")]
  
  out
}

```
```{r,eval=FALSE}
# HMA function is too long, it is hided. 

starttime=proc.time()[3]

HMAout <- tryCatch(HMA_FINAL(dat,X), error = function(x) {
    print("Error with HMA");
    warnings();
    data.frame(    est=rep(NA,6),
                   lb=rep(NA,6),
                   ub=rep(NA,6),
                   se=rep(NA,6),
                   p=rep(NA,6)
    )
  }
)
elapsed=proc.time()[3]-starttime
save(elapsed,HMAout, file="HMA_S1.Rdata")
  
```


### 2.4 TS-NLME and U-NLME

The TS-NLME uses the same dataset structure as the TS-NLS and TS-HMA. We can either use the results from the TS-NLS as starting values or fit the model all over again using broadly chosen starting values.

U-NLME use the fitted value in TS-NLME as starting values, include the linear relations with covariate in the iterated fitting.
```{r,eval=FALSE,echo=FALSE}
#TS-NLME function
nlme_sep_FINAL <- function(myfit){
  # Extract EB estimates of NO parameters (State I)
  ranefdata         <-ranef(myfit,augFrame=T)
  ranefdata$Ca      <-ranefdata$Ca+fixef(myfit)[1]
  ranefdata$logCaw  <-ranefdata$logCaw+fixef(myfit)[2]
  ranefdata$logDaw  <-ranefdata$logDaw+fixef(myfit)[3]
  # Simple linear regressions (Stage II)
  fit_Ca     <- lm(Ca~X,     data=ranefdata)
  fit_logCaw <- lm(logCaw~X, data=ranefdata)
  fit_logDaw <- lm(logDaw~X, data=ranefdata)
  
  # output
  out <- matrix(NA,nrow=6,ncol=5)
  rownames(out) <- c("Ca.(Intercept)","Ca.X","logCaw.(Intercept)","logCaw.X","logDaw.(Intercept)","logDaw.X")
  colnames(out) <- c("est","lb","ub","se","p")
  out <- as.data.frame(out)
  
  out[c("Ca.(Intercept)","Ca.X"),c("est","se","p")] <- summary(fit_Ca)$coef[,c("Estimate","Std. Error","Pr(>|t|)")]
  out[c("logCaw.(Intercept)","logCaw.X"),c("est","se","p")] <- summary(fit_logCaw)$coef[,c("Estimate","Std. Error","Pr(>|t|)")]
  out[c("logDaw.(Intercept)","logDaw.X"),c("est","se","p")] <- summary(fit_logDaw)$coef[,c("Estimate","Std. Error","Pr(>|t|)")]
  
  out[c("Ca.(Intercept)","Ca.X"),c("lb","ub")] <- confint(fit_Ca)[,c("2.5 %","97.5 %")]
  out[c("logCaw.(Intercept)","logCaw.X"),c("lb","ub")] <- confint(fit_logCaw)[,c("2.5 %","97.5 %")]
  out[c("logDaw.(Intercept)","logDaw.X"),c("lb","ub")] <- confint(fit_logDaw)[,c("2.5 %","97.5 %")]
  
  out
}
#Fit TS-NLME
starttime=proc.time()[3]
    
    
    
print("nlmeSepStart")
fitD.nlme <- tryCatch(nlme(nlsList(logeno ~ 
                          log(exp(logCaw) + (Ca-exp(logCaw))*exp(-exp(logDaw)/flow))|id,
                          data=dat, start=list(Ca = 1.5, logCaw =4 , logDaw = 3),
                          control=list(tolerance=0.001)), 
                          random= pdDiag(Ca+logCaw+logDaw ~ 1),
                          verbose=FALSE, 
                          control=list(tolerance = 0.1,maxIter=100,pnlsTol=1e-4,msVerbose=TRUE)), 
                      error = function(x) {
                                 print("Error with nlmeSepstart");
                                 nlmeSepout<-data.frame(    est=rep(NA,6),
                                                            lb=rep(NA,6),
                                                            ub=rep(NA,6),
                                                            se=rep(NA,6),
                                                            p=rep(NA,6)
                                 )
                               }
    )
    
    
print("nlmeSepStart_u")
fitU.nlme <- tryCatch(update(fitD.nlme,random=pdLogChol(Ca+logCaw+logDaw ~ 1),
                             verbose=FALSE), error = function(x) {
      print("Error with nlmeSimStart");
      nlmeSimout<-data.frame(    est=rep(NA,6),
                                 lb=rep(NA,6),
                                 ub=rep(NA,6),
                                 se=rep(NA,6),
                                 p=rep(NA,6)
      )
    }
    )
    
# out put TS-NLME results
nlmeSepout <- tryCatch(nlme_sep_FINAL(fitU.nlme), error = function(x) {
      print("Error with nlmeSep");
      nlmeSepout<-data.frame(est=rep(NA,6),
                             lb=rep(NA,6),
                             ub=rep(NA,6),
                             se=rep(NA,6),
                             p=rep(NA,6)
      )
    }
    )
#U-NLME function
nlme_sim_FINAL <- function(myfit){
  startFix <- fixef(myfit)
  # nlmeSim
  fit4.nlme <- update(myfit,
                      fixed = list(Ca ~ X, logCaw ~ X, logDaw ~ X),
                      start = c(startFix[1],0,startFix[2],0,startFix[3],0),
                      control=list(tolerance=0.01)
  )
  Ttable <- summary(fit4.nlme)$tTable
  lb_est_ub <- intervals(fit4.nlme)$fixed
  out <- data.frame(    est=Ttable[,"Value"],
                        lb=lb_est_ub[,"lower"],
                        ub=lb_est_ub[,"upper"],
                        se=Ttable[,"Std.Error"],
                        p=Ttable[,"p-value"]
  )
  out
  # FOR NLME_sim, you likely want to save additional output for more direct comparison with JAGS
  # like the VAR/COV of random effects, SD error
}

# output U-NLME results
nlmeSimout <- tryCatch(nlme_sim_FINAL(fitU.nlme), error = function(x) {
      print("Error with nlmeSim");
      nlmeSimout<-data.frame(    est=rep(NA,6),
                                 lb=rep(NA,6),
                                 ub=rep(NA,6),
                                 se=rep(NA,6),
                                 p=rep(NA,6)
      )
    }
    )

    
save(nlmeSepout,nlmeSimout, file="NLME_S1.Rdata")    


```

## 3. Model comparisom

Here we show the exmaple code for comparing results generated via each model in a box plot with means and confidence intervals/credible sets.

```{r,eval=FALSE}
load("UHB_S1.Rdata")
load("NLS_S1.Rdata")
load("HMA_S1.Rdata")
load("NLME_S1.Rdata")
library(latex2exp)
JagsXsummary<-result$summary[c("alpha_Ca","alpha_logCaw","alpha_logDaw",
                 "beta_Ca","beta_logCaw","beta_logDaw"),c("mean","2.5%","97.5%")]
CIbeta_Ca<-as.data.frame(
  rbind(as.matrix(NLSout["Ca.X",c("NLS_est","NLS_lb","NLS_ub")]),
        as.matrix(HMAout["Ca.X",c("est","lb","ub")]),
        as.matrix(nlmeSepout["Ca.X",c("est","lb","ub")]),
        as.matrix(nlmeSimout["Ca.X",c("est","lb","ub")]),
        JagsXsummary["beta_Ca",]
        
        ))
CIbeta_Ca$model<-c("01.TS-NLS","02.TS-HMA","03.TS-NLME","04.U-NLME","05.U-HB")
colnames(CIbeta_Ca)<-c("Estimation","lb","ub","model")
library(plotrix)
plotCI(x=c(1:5),y=CIbeta_Ca$Estimation,err="y",ui=CIbeta_Ca$ub,li=CIbeta_Ca$lb,ylab="Estimation",xlab="models",ylim=c(-1,1),xaxt = "n",main=TeX('Determinent of C_{A}'))
axis(side=1,at=c(1:5),labels=XlabCHS,tick=FALSE,cex.axis=0.8)
abline(h=0)


```




